A comparison of design-based and model-based approaches for spatial data.

### 2021-10-12 inst/manuscript/manuscript.Rmd
updates to section 1 and 2

There are two general approaches for using data to make statistical inferences about a population: design-based approaches and model-based approaches. When data cannot be obtained for all units in a population (population units), data on a subset of the population units is collected in a  sample. In the design-based approach, inferences about the underlying population are informed from a probabilistic process in which population units are selected to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process that generated the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of general advantages [@hansen1983evaluation].

@@

__Tony O.__: Should this paragraph address that spatial information can be incorporated in the design stage or in the analysis stage (or both). In general, it's not clear whether we are referring to site selection process or the estimation process

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Thereafter, several comparisons between design-based and model-based for spatial data have been considered, but they tend to compare design-based approaches that ignore spatial locations to model-based approaches [@brus1997random; @verhoef2002sampling; @verhoef2008spatial]. @cooper2006sampling review the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g. @cicchitelli2012model, @chan2020bayesian for a Bayesian approach, and @sterba2009alternative). More recent overviews include @brus2020statistical and @wang2012review, but no numerical comparison has been made between design-based approaches that incorporate spatial locations and model-based approaches.

<!-- Outline for the rest of the paper -->

__Lisa M.__: Add paragraph describing contribution of manuscript.

The rest of this paper is organized as follows. In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach. In Section \ref{sec:numstudy}, we use simulated and real data to study the the behavior of both approaches. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

## Comparing Design-Based vs. Model-Based

The design-based approach assumes the population is fixed. Randomness is incorporated in the selection of population units according to a sampling design. A sampling design assigns a positive probability of inclusion in the sample (inclusion probability) to each population unit. Some examples of commonly used sampling designs include simple random sampling, stratified random sample, and cluster
sampling, which we refer to as Independent Random Sampling (IRS) survey designs. The goal is to use the sampling design and the sampled data to estimate population parameters like means and totals. These population parameters are traditionally assumed to be fixed but unknown.

Treating the data as fixed and incorporating randomness through the sampling design (top row of Figure \ref{fig:fig1} ((cite Brus 2021 here since our figure is similar?))) yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments. Means and totals, for example, are asymptotically normally distributed by the Central Limit Theorem. If we repeatedly sample the surface, then 95% of all 95% confidence intervals constructed from a procedure with appropriate coverage will contain the true, fixed mean. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

__Jay VH__: I think it is important to stress that the limiting distribution is over all possible randomizations, constrained by whatever design is used.

__Jay VH__: quantity is vague. We should stick with variables, or realized variables (we might also call these values, but we should define and establish a consistent terminology early on.) __Matt H__: I think, though this comment is for this paragraph, we should establish the terminology earlier.

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process and need not be incorporated through random sampling (bottom row of Figure \ref{fig:fig1}). Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty.

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals to be more precise. If we repeatedly generate the response values from a fixed spatial process and obtain a sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

@@

# Application

>>>>

There are two general approaches for using data to make statistical inferences about a population: design-based and model-based. When data cannot be obtained for all units in a population (population units), data on a subset of the population units is collected and called a sample. In the design-based approach, inferences about the underlying population are informed from a probabilistic process in which population units are selected to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process that generated the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of general advantages [@hansen1983evaluation].

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Thereafter, several comparisons between design-based and model-based for spatial data have been considered [@brus1997random; @verhoef2002sampling; @verhoef2008spatial]. @cooper2006sampling review the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g. @cicchitelli2012model, @chan2020bayesian for a Bayesian approach, and @sterba2009alternative). More recent overviews include @brus2020statistical and @wang2012review.

<!-- Paragraph about the contribution of the manuscript and outlining the rest -->
Though comparisons between design-based and model-based approaches to spatial data have been studied, no numerical comparison has been made between design-based approaches that incorporate spatial locations and model-based approaches. In this manuscript, we compare design-based approaches that incorporate spatial locations to model-based approaches for spatial data. The rest of the manuscript is organized as follows. In Section \ref{sec:background}, we compare sampling and estimation procedures between the design-based approach and the model-based approach for spatial data. In Section \ref{sec:numstudy}, we use a simulation approach to study the behavior and performance of both approaches. In Section \ref{application}, we use both approaches to analyze real data. And in Section \ref{sec:discussion}, we end with a discussion and provide directions for future research.

@@

## Comparing Design-Based and Model-Based Approaches

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of units in a sampling frame according to a sampling design. A sampling frame is the set of all units available to be sampled. A sampling design assigns a positive probability of inclusion (inclusion probability) to each unit in the sampling frame. Some examples of commonly used sampling designs include simple random sampling, stratified random sample, and cluster
sampling. These sampling designs tend to select units from the sampling frame independently of other units, so we call them "Independent Random Sampling" (IRS) designs. The design-based approach combines the randomness of the sampling design and the data collected via the sample to estimate parameters (e.g. means and totals) of a population. Generally, these population parameters are assumed to be fixed, unknown constants.

Treating the data as fixed and incorporating randomness through the sampling design yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments that incorporate all possible randomizations of sampling units selected via the sampling design. Means and totals, for example, are asymptotically Gaussian by the Central Limit Theorem (under some assumptions). If we repeatedly sample the surface, then 95% of all 95% confidence intervals constructed from a procedure with appropriate coverage will contain the true, fixed mean. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

The model-based approach assumes the data are a random realization of a data-generating process. Randomness is often incorporated through distributional assumptions on this process and need not be incorporated through random sampling (bottom row of Figure \ref{fig:fig1}). Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived leveraging assumptions of the data generating process. If we repeatedly generate the response values from a fixed spatial process and obtained a sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data. A visual comparison of the design-based and model-based assumptinos is provided in Figure \ref{fig:fig1} (@brus2020statistical provides a similar figure).

@@

# Application {#application}

### 2021-10-12 inst/manuscript/manuscript.tex
updates to section 1 and 2

There are two general approaches for using data to make statistical
inferences about a population: design-based approaches and model-based
approaches. When data cannot be obtained for all units in a population
(population units), data on a subset of the population units is
collected in a sample. In the design-based approach, inferences about
the underlying population are informed from a probabilistic process in
which population units are selected to be in the sample. Alternatively,
in the model-based approach, inferences are made from specific
assumptions about the underlying process that generated the data. Each
paradigm has a deep historical context (Sterba, 2009) and its own set of
general advantages (Hansen et al., 1983).

\textbf{Tony O.}: Should this paragraph address that spatial information
can be incorporated in the design stage or in the analysis stage (or
both). In general, it's not clear whether we are referring to site
selection process or the estimation process

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
locations to model-based approaches (Brus and De Gruijter, 1997; Ver
Hoef, 2002, 2008). Cooper (2006) review the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design and model-based principles (see
e.g.~Cicchitelli and Montanari (2012), Chan-Golston et al. (2020) for a
Bayesian approach, and Sterba (2009)). More recent overviews include
Brus (2020) and Wang et al. (2012), but no numerical comparison has been
made between design-based approaches that incorporate spatial locations
and model-based approaches.

\textbf{Lisa M.}: Add paragraph describing contribution of manuscript.

The rest of this paper is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach. In
Section \ref{sec:numstudy}, we use simulated and real data to study the
the behavior of both approaches. And in Section \ref{sec:discussion}, we
end with a discussion and provide directions for future research.

@@

\hypertarget{comparing-design-based-vs.-model-based}{%
\subsection{Comparing Design-Based
vs.~Model-Based}\label{comparing-design-based-vs.-model-based}}

The design-based approach assumes the population is fixed. Randomness is
incorporated in the selection of population units according to a
sampling design. A sampling design assigns a positive probability of
inclusion in the sample (inclusion probability) to each population unit.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sample, and cluster sampling, which we refer
to as Independent Random Sampling (IRS) survey designs. The goal is to
use the sampling design and the sampled data to estimate population
parameters like means and totals. These population parameters are
traditionally assumed to be fixed but unknown.

Treating the data as fixed and incorporating randomness through the
sampling design (top row of Figure \ref{fig:fig1} ((cite Brus 2021 here
since our figure is similar?))) yields estimators having very few other
assumptions. Confidence intervals for these types of estimators are
typically derived using limiting arguments. Means and totals, for
example, are asymptotically normally distributed by the Central Limit
Theorem. If we repeatedly sample the surface, then 95\% of all 95\%
confidence intervals constructed from a procedure with appropriate
coverage will contain the true, fixed mean. Särndal et al. (2003) and
Lohr (2009) provide thorough reviews of the design-based approach.

\textbf{Jay VH}: I think it is important to stress that the limiting
distribution is over all possible randomizations, constrained by
whatever design is used.

\textbf{Jay VH}: quantity is vague. We should stick with variables, or
realized variables (we might also call these values, but we should
define and establish a consistent terminology early on.) \textbf{Matt
H}: I think, though this comment is for this paragraph, we should
establish the terminology earlier.

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is often incorporated through
distributional assumptions on this process and need not be incorporated
through random sampling (bottom row of Figure \ref{fig:fig1}). Instead
of estimating fixed but unknown parameters (as in the design-based
approach), the goal of model-based inference in the spatial context is
often \emph{prediction} of an unknown quantity. For example, suppose the
realized mean of all population units is the quantity of interest.
Instead of \emph{estimating} a fixed unknown mean, we are
\emph{predicting} the value of the mean, a random variable. We know that
if we sampled all population units, we would have an exact prediction
for the mean of our one realized process, without any uncertainty.

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals to be more
precise. If we repeatedly generate the response values from a fixed
spatial process and obtain a sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data.

>>>>

There are two general approaches for using data to make statistical
inferences about a population: design-based and model-based. When data
cannot be obtained for all units in a population (population units),
data on a subset of the population units is collected and called a
sample. In the design-based approach, inferences about the underlying
population are informed from a probabilistic process in which population
units are selected to be in the sample. Alternatively, in the
model-based approach, inferences are made from specific assumptions
about the underlying process that generated the data. Each paradigm has
a deep historical context (Sterba, 2009) and its own set of general
advantages (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered (Brus
and De Gruijter, 1997; Ver Hoef, 2002, 2008). Cooper (2006) review the
two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design and
model-based principles (see e.g.~Cicchitelli and Montanari (2012),
Chan-Golston et al. (2020) for a Bayesian approach, and Sterba (2009)).
More recent overviews include Brus (2020) and Wang et al. (2012).

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. The rest of the manuscript is organized as follows. In
Section \ref{sec:background}, we compare sampling and estimation
procedures between the design-based approach and the model-based
approach for spatial data. In Section \ref{sec:numstudy}, we use a
simulation approach to study the behavior and performance of both
approaches. In Section \ref{application}, we use both approaches to
analyze real data. And in Section \ref{sec:discussion}, we end with a
discussion and provide directions for future research.

@@

\hypertarget{comparing-design-based-and-model-based-approaches}{%
\subsection{Comparing Design-Based and Model-Based
Approaches}\label{comparing-design-based-and-model-based-approaches}}

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sample, and cluster sampling. These sampling
designs tend to select units from the sampling frame independently of
other units, so we call them ``Independent Random Sampling'' (IRS)
designs. The design-based approach combines the randomness of the
sampling design and the data collected via the sample to estimate
parameters (e.g.~means and totals) of a population. Generally, these
population parameters are assumed to be fixed, unknown constants.

Treating the data as fixed and incorporating randomness through the
sampling design yields estimators having very few other assumptions.
Confidence intervals for these types of estimators are typically derived
using limiting arguments that incorporate all possible randomizations of
sampling units selected via the sampling design. Means and totals, for
example, are asymptotically Gaussian by the Central Limit Theorem (under
some assumptions). If we repeatedly sample the surface, then 95\% of all
95\% confidence intervals constructed from a procedure with appropriate
coverage will contain the true, fixed mean. Särndal et al. (2003) and
Lohr (2009) provide thorough reviews of the design-based approach.

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is often incorporated through
distributional assumptions on this process and need not be incorporated
through random sampling (bottom row of Figure \ref{fig:fig1}). Instead
of estimating fixed but unknown parameters (as in the design-based
approach), the goal of model-based inference in the spatial context is
often to predict a realized variable, or value. For example, suppose the
realized mean of all population units is the value of interest. Instead
of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the
value of the mean, a random variable. We know that if we sampled all
population units (and not any units outside the population), we would be
able to exactly predict (without any uncertainty) the mean of our one
realized process. Assuming the data are realized from a specific
data-generating process yields predictions that are linked to
distributional assumptions. These distributional assumptions are used to
derive prediction intervals. If we repeatedly generate the response
values from a fixed spatial process and obtained a sample, then 95\% of
all 95\% prediction intervals constructed from a procedure with
appropriate coverage will contain their respective realized means.
Cressie (1993) and Schabenberger and Gotway (2017) provide reviews of
model-based approaches for spatial data. A visual comparison of the
design-based and model-based assumptinos is provided in Figure
\ref{fig:fig1} (Brus (2020) provides a similar figure).
